---
title: "Infectious_disease_modelling_case scenario 1"
author: "Finta Francis"
format: html
editor: visual
---
# **Modeling Malaria Transmission Dynamics in Nigeria Using a Vector–Host Framework**

## **Abstract**

Malaria remains one of the most significant public health challenges in Nigeria, contributing substantially to global morbidity and mortality. This study develops a deterministic vector–host transmission model to simulate malaria dynamics and evaluate the impact of intervention strategies. Using a compartmental SEIR (human)–SEI (mosquito) framework, we simulate three intervention scenarios: increased insecticide-treated net (ITN) coverage (75%), indoor spraying access (50%), and treatment access (45%). The model is implemented in R using the deSolve package. Results highlight the relative effectiveness of combined interventions in reducing infection prevalence.

---

# 1. Introduction

Malaria is a life-threatening parasitic disease transmitted through bites of infected *Anopheles* mosquitoes. According to the World Health Organization, sub-Saharan Africa accounts for over 90% of global malaria cases and deaths, with Nigeria alone contributing a significant proportion.

### Why Modeling is Needed

* Malaria is **endemic**, not episodic
* Transmission depends on **vector dynamics**
* Interventions interact in **nonlinear ways**
* Field experiments are costly → models provide insight

Mathematical models allow us to:

* Predict disease spread
* Evaluate interventions before implementation
* Identify key transmission drivers

---

# 📖 2. Literature Review & Parameter Justification

##  Biting Rate (a)

* Empirical studies in Nigeria report:

  * ~6–17 bites/person/night (field studies)
* Scaled in models to:

  * **0.2–0.5 per day**

##  Transmission Probabilities

* Mosquito → human (**b**): 0.1–0.5
* Human → mosquito (**c**): 0.1–0.3

##  Incubation Period

* Humans: ~10–14 days
* Mosquitoes: ~10 days

##  Mosquito Mortality

* Lifespan: 7–14 days

---

# 3. Model Formulation

We use a **vector-host SEIR-SEI model**.

## Human Population:

* Susceptible (Sₕ)
* Exposed (Eₕ)
* Infected (Iₕ)
* Recovered (Rₕ)

## Mosquito Population:

* Susceptible (Sₘ)
* Exposed (Eₘ)
* Infected (Iₘ)

---

## Transmission Dynamics

* Infection from mosquito to human:

  * Depends on biting rate and infected mosquitoes
* Infection from human to mosquito:

  * Depends on infected humans

---

# 4. Model Implementation in R

##  Load Package

```{r}
install.packages("deSolve")
library(deSolve)
```

## Model Equations

```{r, hunk 2 }
malaria_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    
    Nh <- Sh + Eh + Ih + Rh
    Nm <- Sm + Em + Im
    
    # Adjusted biting rate (interventions affect this)
    a_eff <- a * (1 - ITN_effect) * (1 - spray_effect)
    
    lambda_h <- a_eff * b * Im / Nm
    lambda_m <- a_eff * c * Ih / Nh
    
    dSh <- mu_h * Nh - lambda_h * Sh - mu_h * Sh
    dEh <- lambda_h * Sh - sigma_h * Eh - mu_h * Eh
    dIh <- sigma_h * Eh - gamma_h * Ih - mu_h * Ih
    dRh <- gamma_h * Ih - mu_h * Rh
    
    dSm <- lambda_m * Sm 
    dEm <- lambda_m * Sm - sigma_m * Em
    dIm <- sigma_m * Em - mu_m * Im
    
    list(c(dSh, dEh, dIh, dRh, dSm, dEm, dIm))
  })
}
```

## Parameters

```{r, chunk3}
base_parameters <- c(
  a = 0.3,          # biting rate
  b = 0.3,          # mosquito → human transmission
  c = 0.2,          # human → mosquito transmission
  
  sigma_h = 1/10,   # human incubation
  sigma_m = 1/10,   # mosquito incubation
  gamma_h = 1/14,   # recovery rate
  
  mu_h = 1/(60*365),# human death rate
  mu_m = 1/10,      # mosquito death rate
  
  ITN_effect = 0,
  spray_effect = 0
)
```

## Initial Conditions

```{r, chunk 4}
state <- c(
  Sh = 990, Eh = 5, Ih = 5, Rh = 0,
  Sm = 10000, Em = 100, Im = 100
)
```

##  Simulation Time

```{r, chunk 5}
times <- seq(0, 365, by = 1)

params_base <- base_parameters
```

# 5. Intervention Scenarios

We test three hypotheses:

---

## Scenario 1: 75% ITN Coverage

Effect:

* Reduces biting rate

```{r, chunk 6}
params_itn <- base_parameters
params_itn["ITN_effect"] <- 0.75 * 0.6
```

## Scenario 2: 50% Indoor Spraying

```{r, chunk 7}
params_spray <- base_parameters
params_spray["spray_effect"] <- 0.5 * 0.5
```

## Scenario 3: 45% Treatment Coverage

Effect:

* Increases recovery rate

```{r, chunk 8}
params_treat <- base_parameters
params_treat["gamma_h"] <- 1/10
```

# Run Simulations

```{r, chunk 9}
out_base <- ode(state, times, malaria_model, params_base)
out_itn <- ode(state, times, malaria_model, params_itn)
out_spray <- ode(state, times, malaria_model, params_spray)
out_treat <- ode(state, times, malaria_model, params_treat)

out_base <- as.data.frame(out_base)
out_itn <- as.data.frame(out_itn)
out_spray <- as.data.frame(out_spray)
out_treat <- as.data.frame(out_treat)

```

## ** SEIR plot

```{r, plot chunk 1}
matplot(out_base$time,
        cbind(out_base$Sh, out_base$Eh, out_base$Ih, out_base$Rh),
        type = "l",
        lty = 1,
        col = c("blue", "orange", "red", "green"),
        xlab = "Time (days)",
        ylab = "Population")

legend("right",
       legend = c("S", "E", "I", "R"),
       col = c("blue", "orange", "red", "green"),
       lty = 1)
```


## ** Mosquito Dynamics

```{r, chunk 10}
plot(out$time, out$Sm, type="l",
     xlab="Time (days)", ylab="Mosquito Population",
     lty=1)

lines(out$time, out$Em, lty=2)
lines(out$time, out$Im, lty=3)

legend("right",
       legend=c("Susceptible Mosq","Exposed Mosq","Infected Mosq"),
       lty=1:3)
```
## ** Comparison scenarios

```{r, chunk12}
plot(out_base$time, out_base$Ih, type="l", col="black",
     xlab="Time", ylab="Infected Humans")

lines(out_itn$time, out_itn$Ih, col="blue")
lines(out_spray$time, out_spray$Ih, col="green")
lines(out_treat$time, out_treat$Ih, col="red")

legend("topright",
       legend=c("Baseline","ITN","Spray","Treatment"),
       col=c("black","blue","green","red"),
       lty=1)
```

