---
title: "Infectious disease modelling_SI model"
author: "Finta Francis"
format: html
editor: visual
---

## *** SI MODEL STRUCTURE

## ------We only track --------

S (Susceptible humans)
I (Infected humans)

Total population:

N=S+I( Here we have only two groups, those who are susceptible, and those who are infected )

Model equations

dS/ dt = μN−βSI−μS
( Here we are assuming the population to be constant, births = deaths, so we added μ * N , to the population , who is suceptible, μ is the mortality rate  )

 dI / dt = βSI−μI

Where:

β = transmission rate from susceptible to infectiousness . In SI model, we assume that once S moves to I, they remain in the infectious compartment.

μ = natural death rate (keeps population stable)

```{r, chunk 1}
library(deSolve)
```

You can add options to executable code like this

```{r, model functions}
si_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    
    N <- S + I
    
    # Force of infection
    lambda <- beta * I
    
    dS <- mu * N - lambda * S - mu * S
    dI <- lambda * S - mu * I
    
    list(c(dS, dI))
  })
}
```
## ** Parameters 

Here we define the paprameters in the above equation

```{r, parameters}
parameters <- c(
  beta = 0.0005,   # transmission rate
  mu = 1/(60*365)  # death rate
)
```

## ** Initial conditions ** ##

```{r, initials}

## Here we assume that in the total population of 10000, 100 people got infected, so I = 100, S = N - I , 

state <- c(
  I = 10,
  S = 990
)
```

## ** Time and simulations ** ## 

```{r, time and simukations }
times <- seq(0, 365, by = 1)

## simulations 

out <- ode(y = state, times = times, func = si_model, parms = parameters)
out <- as.data.frame(out)
```
##  ** Plots ** ###

```{r, plots}
plot(out$time, out$S, type = "l",
     col = "blue",
     xlab = "Time (days)",
     ylab = "Population")

lines(out$time, out$I, col = "red")

legend("right",
       legend = c("Susceptible", "Infected"),
       col = c("blue", "red"),
       lty = 1)
```

In this model, the disease free eullibrium is maintained as susceptible increases, and the infectiousness decreases.